# NOAA Storm Data Analysis
## *Storm event types and damages*

-------

### Summary
This analysis addresses the following questions:

1. Across the United States, which types of events (as indicated in the <code>EVTYPE</code> variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

--------

### Data Processing

*STEP 1:* Read and cache the data
```{r data_processing, cache=TRUE}
data <- read.csv("repdata_data_StormData.csv.bz2")
```

*STEP 2:* Subset Data
A summary of the NOAA Storm Database tracks a myriad of weather phenomen and their impact on our civilization. For the purposes of this analysis, we will only consider the following variables:
```{r summary}
columns <- c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
subdata <- data[columns]
summary(subdata)
```

*STEP 3:* Verify Data Types
As we can see below, the data types are ok for use in this analysis.
```{r data_types}
subdata_types <- vector()
for(i in 1:ncol(subdata)){
    subdata_types[i] <- class(subdata[,i])
}
vars <- names(subdata)
subdata_types <- data.frame(vars, subdata_types)
names(subdata_types) <- c("variable", "data_type")
subdata_types
```

*STEP 4:* Process Data
Per the data logging instructions, all monetary values are store in 2 values: *PROPDMG*/*PROPDMGEXP* and C*CROPDMG*/*CROPDMGEXP*. To obtain the real monetary value as a numeric, we must 

**4.1)** Convert the value in *DMGEXP* to numeric. The codebook only describes *B* (billions), *M* (millions), and *K* (thousands). However the data shows the following values:

```{r}
unique(subdata$PROPDMGEXP)
```

The values *+*,*-*, *""*, and *?* are not defined and will be interpreted as 1 to preserve any raw value shown in the corresponding *PROPDMG* column. The numeric values are also not defined in the code book and will be defined as *1e+01*, *1e+02*, etc...
 
```{r process_data}
subdata$PROPDMGEXP <- as.vector(subdata$PROPDMGEXP)
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "H"] <- 100
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "h"] <- 100
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "K"] <- 1000
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "m"] <- 1e+06
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "M"] <- 1e+06
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "B"] <- 1e+09
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "1"] <- 1e+01
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "2"] <- 1e+02
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "3"] <- 1e+03
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "4"] <- 1e+04
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "5"] <- 1e+05
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "6"] <- 1e+06
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "7"] <- 1e+07
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "8"] <- 1e+08
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "+"] <- 1
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "-"] <- 1
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "?"] <- 1
subdata$PROPDMGEXP[subdata$PROPDMGEXP == "0"] <- 1
subdata$PROPDMGEXP[subdata$PROPDMGEXP == ""] <- 1

subdata$CROPDMGEXP <- as.vector(subdata$CROPDMGEXP)
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "H"] <- 100
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "h"] <- 100
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "K"] <- 1000
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "m"] <- 1e+06
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "M"] <- 1e+06
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "B"] <- 1e+09
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "1"] <- 1e+01
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "2"] <- 1e+02
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "3"] <- 1e+03
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "4"] <- 1e+04
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "5"] <- 1e+05
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "6"] <- 1e+06
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "7"] <- 1e+07
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "8"] <- 1e+08
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "+"] <- 1
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "-"] <- 1
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "?"] <- 1
subdata$CROPDMGEXP[subdata$CROPDMGEXP == "0"] <- 1
subdata$CROPDMGEXP[subdata$CROPDMGEXP == ""] <- 1

subdata$CROPDMGEXP <- as.numeric(subdata$CROPDMGEXP)
subdata$PROPDMGEXP <- as.numeric(subdata$PROPDMGEXP)
head(subdata[subdata$PROPDMG > 0])
head(subdata[subdata$CROPDMG > 0])
```
 
**4.2** Multiply the *DMG* by the value described in *DMGEXP*.
```{r createdmgvalue}
subdata$PROPDMG <- subdata$PROPDMG * subdata$PROPDMGEXP
head(subdata[subdata$PROPDMG > 0,])
head(subdata[subdata$CROPDMG > 0,])
```

*STEP 5:* Problems with *EVTYPE* (Event Type)
The codebook only defines *48 different Event Types*.  
```{r eventtypes}
counteventtypes <- length(unique(subdata$EVTYPE))
```
However, the data shows `counteventtypes`.

Since we have no frame of reference for why there are so many more event types than defined, we will not change the event types in any way at this point. There is no way to know what the data collector was thinking at the time of recording. For the purpose of this analysis, we will assume the codebook is not up to date.

*STEP 6:* Figuring out ppopulation harm
To determine which events have the most harm in relation to population health, we will add the number of fatalities and injuries together by event and store it in a data frame.

```{r populationhealth}
require(plyr)
pophealth <- ddply(subdata, .(levels(EVTYPE)), summarise, popharm=FATALITIES+INJURIES)
pophealth <- arrange(pophealth, desc(popharm))
tpopharm <- pophealth[1:25, ]
names(tpopharm) <- c("EVTYPE", "Population_Harm")
tpopharm$EVTYPE <- factor(tpopharm$EVTYPE)
tpopharm
```

*STEP 7:* Figuring out economic consequences
To determine which events have the most greatest economic consequences, we will 

 - multiply PROPDMG by PROPDMGEXP
 - multiply CROPDMG by CROPDMGEXP
 - sum the results of the previous two operations by EVTYPE
 - store the results in a dataframe
 
```{r econconsequences}
econsequences <- ddply(subdata, .(levels(EVTYPE)), summarise, econ=((PROPDMG*PROPDMGEXP)+(CROPDMG*CROPDMGEXP)))
econsequences <- arrange(econsequences, desc(econ))
tecon <- econsequences[1:25, ]
names(tecon) <- c("EVTYPE", "Economic_Consequences")
tecon$EVTYPE <- factor(tecon$EVTYPE)
tecon
```


### RESULTS

1. Across the United States, which types of events (as indicated in the <code>EVTYPE</code> variable) are most harmful with respect to population health?

```{r mostharmful}
p1 <- ggplot(tpopharm, aes(EVTYPE, Population_Harm))
p1 <- p1 + geom_bar(stat="identity")
p1 <- p1 + theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))
p1
```

From the chart above, it is clear that the most harmful events to population health are:

```{r r1}
tpopharm[1:2,]
```

2. Across the United States, which types of events have the greatest economic consequences?

```{r mostecon}
p1 <- ggplot(tecon, aes(EVTYPE, Economic_Consequences))
p1 <- p1 + geom_bar(stat="identity")
p1 <- p1 + theme(axis.text.x=element_text(angle=45,hjust=1,vjust=1))
p1
```





- 